Fast Sorting on Modern Computers

### 2005-07-19 docs/tech_report/radix.tex
added aligned cache radix sort to discussion. Made minor fixes

The steps to sort by radix are simple. First a radix is chosen, based on the
size of the part to be sorted. For 32 bits integers, a byte is used. This means
the radix is 256, as there are 256 possible byte values. An array of 256
counters is created to count the number of keys with a particular least
significant byte. Then the counters are summed in turn to store the cumulative
count of keys that belong before them in the array. Now each byte can be put in
place by accessing the counter indexed by its least significant byte.

These are written into an auxiliary array, which is written back before the next
iteration using the second least significant byte. Since every key is processed
in order, when the next most significant byte is sorted, the order of the less
significant bytes are maintained, so that once the most significant byte is
sorted, the entire array is.

This type of radixsort is called \n{Least Significant Digit} radixsort.
Other implementations are possible, and several improvements also exist, but as
radixsort is included here for comparative purposes, these are not implemented.
They are, however, discussed in Section \ref{radix future work}.

@@

The first improvement is to reduce the number of runs across the array per
step.  In the \n{base radixsort}, presented above, there is one step to count,
one to copy, and one to write back (which accesses both arrays). LaMarca
suggests counting for the next step in the previous count. In fact, all the
counting can be done in a single pass at the start, instead of four passes at
various stages, decreasing instruction count and cache misses. The downside is
the fourfold increase in memory required for the arrays of counts, with no
benefits.

The step to copy the array back can also be removed. In mergesort, copying back
to the source array was changed to copying back and forth between the source and 
auxiliary arrays. The same could be potentially done with radixsort, without any
increase in complexity. The number of steps is likely to be even, and is known
at compile time, so the number of times the copy step is done could be reduced
to zero. If the number of steps is odd, which is unlikely when sorting
integers, then one copy at the end would be required. This is not needed in the
implementation presented here.

@@

@@

\subsection{Expected Performance}
Radixsort is expected to have very good performance, compared to comparison
based sorts. This is due to that fact that radixsort is $O(N)$; for an array as
small as 4096 records, it involves 12 times fewer steps than the $O(NlogN)$
sorts.

The cache performance of base radixsort is not expected to be good, however.
Quicksort and mergesort, for example, have very good temporal locality. Keys are
used repeatedly before being swapped out of the cache. In radixsort, however,
every key is used just once each time it is in the cache. When the array is larger than
the size of the cache, every key access will be a miss as a result, and every
new byte being sorted involves four passes over the array. In addition, if the
auxiliary array is pathologically aligned, then the copy back into the array
can result in thrashing.

@@

The number of branch prediction misses should be exceptionally low. In
comparison based sorts, a branch due to a comparison can not be predicted except
in exceptional circumstances\footnote{See Section \ref{insertion is
predictable}.}. As a result, only flow control branches occur in radixsort, and
they are all predictable.

>>>>

The steps to sort by radix are simple. First a radix is chosen, based on the
size of the part to be sorted. For 32 bits integers, a byte is used. This means
the radix is 256, as there are 256 possible byte values. An array of 256
counters is created to count the number of keys with a particular least
significant byte. The counters are then summed in turn to store the cumulative
count of keys with a smaller least significant byte. Now each key can be put in
place by accessing the counter indexed by its least significant byte.

These are written into an auxiliary array, which is written back to the initial
array before the next iteration. The process is then repeated using the second
least significant byte.  Since every key is processed in order, when the next
most significant byte is sorted, the order of the less significant bytes are
maintained, so that once the most significant byte is sorted, the entire array
is.

This type of radixsort is called \n{Least Significant Digit} radixsort.
Other implementations are possible, inclduing a \n{Most significant bit}
version, which behaves as a divide and conquer algortithm. These are discussed
in Section \ref{radix future work}.

@@

The first improvement is to reduce the number of runs across the array per
step.  In the \n{base radixsort}, presented above, there is one step to count,
one to copy, and one to write back (which accesses both arrays). LaMarca
suggests counting for the next step in the previous count, which we implement.

In fact, all the counting could be done in a single pass at the start, instead
of four passes at various stages, decreasing instruction count and cache misses.
The downside is the fourfold increase in memory required for the arrays of
counts, with no benefits.

The step to copy the array back can also be removed. In mergesort, copying back
to the source array was changed to copying back and forth between the source and
auxiliary arrays. The same could be done with radixsort, without any increase in
complexity. The number of steps is likely to be even, and is known at compile
time, so the number of times the copy step is done could be reduced to zero. If
the number of steps is odd, which is unlikely when sorting integers, then one
copy at the end would be required. This is not needed in the implementation
presented here.

@@

\subsection{Aligned Cache Radixsort}
In Section \ref{tiled mergesort}, mergesort is improved by aligning its array
with reference to it auxilliary array. Specifically, extra memory is allocated
for the auxilliary array, which is then offset so that the start of the two
arrays are exactly half the length of the cache apart. We further added an
alignment for the level 1 cache to significantly reduce level 1 misses at the
expense of a small number of level 2  misses. We adapt this strategy here, and
call it Aligned Cache Radixsort.

@@

\subsection{Expected Performance}
Radixsort is expected to have very good performance, compared to comparison
based sorts. This is due to that fact that radixsort is $O(N)$; for an array as
small as 4096 records, it involves 12 times fewer steps than the $O(NlogN)$
sorts. For an array as large as 4194304, an $(O(NlogN)$ sort will take 22 times
as many steps to complete. However, it is not expected that radixsort will
perform many times better than quicksort, the reigning champion. A `step' for
quicksort is as simple as a comparison and a possible exchange every few steps.
For radixsort, a step comprises being counted four times and copied four times,
with the counting involving several layers of indirection, and causing a
complete iteration across the array at the start. LaMarca's results
show the instruction count of quicksort and radixsort being in the same order of
magnitude, and we expect the same.

The cache performance of base radixsort is not expected to be good, however.
Quicksort and mergesort have very good temporal locality. Keys are
used repeatedly before being swapped out of the cache. In radixsort, however,
every key is used just once each time it is in the cache. When the array is larger than
the size of the cache, every key access will be a miss as a result, and every
new byte being sorted involves four passes over the array. In addition, if the
auxiliary array is pathologically aligned, then the copy back into the array
can result in thrashing.

The thrashing can also occur in aligned radixsort. Unlike mergesort, we cannot
guarantee that the keys being considered are at the same positions in both
arrays. In fact, we have no idea where is being accessed in the source array, so
it is unlikely that the alignment will have any effect what-so-ever.

@@

The number of branch prediction misses should be exceptionally low. In
comparison based sorts, a branch due to a comparison can not be predicted except
in exceptional circumstances\footnote{See Section \ref{insertion is
predictable}.}. Radixsort has no comparative branches at all. As a result, only
flow control branches occur in radixsort, and they are all predictable.
